On-the-fly Algorithms for GKAT Equivalences

### 2025-01-23 main.tex
edits to section 5

The final piece of the puzzle is to convert any given expression into an \emph{equivalent} state in some symbolic GKAT coalgebra. 
This goal can be achieved by lifting existing constructions like derivatives and Thompson's construction~\cite{schmid_GuardedKleeneAlgebra_2021,smolka_GuardedKleeneAlgebra_2020} to the symbolic setting.
Then the correctness of non-symbolic version can be used to show the correctness of the symbolic construction i.e. we will prove that the lowering as shown in construction~\labelcref{cons:lowering} of these constructions will yield the conventional derivative.
However, several other important properties, like finiteness of derivative coalgebra and the correctness of the Thompson's construction can be established using a symbolic GKAT coalgebra homomorphism from the Thompson's construction to the derivative.

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
In more plain words, a transition is in \(D̂\) if and only if it is derivable by the rules in~\Cref{fig:derivatives-rules}.
These rules are very close to the rules of GKAT derivative by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence, as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis on our correctness argument.
\begin{theoremEnd}{theorem}[Correctness]\label{thm:derivative-correctness}
    The lowering of \(D̂\), denoted \(D\), is exactly the derivative defined by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics generated by the derivative coalgebra, \(⟦e⟧ = ⟦e⟧_{D} = ⟦e⟧_{D̂}\).
\end{theoremEnd}
\begin{proofEnd}
    TODO: unfold the statement.
\end{proofEnd}

@@

Another way to construct a coalgebra from a GKAT expression is via Thompson's construction, we lift the original construction to the symbolic setting.
A useful operation for Thompson's construction is the following guard operation, denoted by \(⟨B|\), where \(B\) is a set of boolean expressions: for a transition function \(δ̂\), a accepting function \(ϵ̂\), and a state \(s\),
\begin{align*}
    ⟨B|ϵ̂(s) & ≜ \{b ∧ c ∣ b ∈ B, c ∈ \hat{\epsilon}(s)\}; \\
    ⟨B|δ̂(s) & ≜ \{(b ∧ c, s', p) ∣ b ∈ B, (c, s', p) ∈ \hat{\delta}(s)\}.
\end{align*}
Besides guarding transition and acceptance with different conditions in if statements and while loops, the guard operator can also be used to simulate uniform continuation. Specifically, we can use \(⟨ϵ̂(s)|δ(s')\) to connecting all the accepting transition of \(s\) to the dynamic \(δ(s')\).

With these definitions in mind, we can define symbolic Thompson's construction inductively as in~\Cref{tab:symb-Thompson-construction}, where we let \((S₁, ϵ̂₁, δ̂₁)\) and \((S₂, ϵ̂₂, δ̂₂)\) to be result of Thompson's construction for \(e₁\) and \(e₂\) respectively.
In this table, \(S₁ + S₂\) denotes the disjoint union of \(S₁\) and \(S₂\), and for any two transition dynamics \(δ₁(s₁): \mathscr{P}(\BExp × S₁ × K)\) and \(δ₂(s₂): \mathscr{P}(\BExp × S₂ × K)\), we can also compose them in parallel as \(δ₁(s₁) + δ₂(s₂)\):
\begin{align*}
    δ₁(s₁) & + δ₂(s₂): \mathscr{P}(\BExp × (S₁ + S₂) × K) \\*
    δ₁(s₁) & + δ₂(s₂) ≜ \\*
        & \{(b, \mathrm{inj}ₗ(s₁'), p) ∣ (b, s₁', p) ∈ δ₁(s₁)\} ∪ \\*
        & \{(b, \mathrm{inj}ᵣ(s₂'), p) ∣ (b, s₂', p) ∈ δ₂(s₂)\},
\end{align*}
where \(\mathrm{inj}ₗ: S₁ → S₁ + S₂\) and \(\mathrm{inj}ᵣ: S₂ → S₁ + S₂\) are the canonical left/right injection of the coproduct.

Our construction deviates from the original construction~\cite{smolka_GuardedKleeneAlgebra_2020} by using a start state \(s^* ∈ S\) instead of a start dynamics (or pseudo-state).
This choice will make the proof of~\Cref{thm:hom-thompson-derivative} slightly easier. 
However, in~\Cref{sec:optimization-implementation}, we will explain that our implementation uses start dynamics instead of start state, to avoid unnecessary lookups and unreachable states.

We would like to explore several desirable theoretical properties of both derivatives and Thompson's construction.
Specifically, the \emph{correctness}, i.e. the semantics of the ``start state'' the both construction have the same preserves the trace semantics of the expression; \emph{finiteness}, i.e. the coalgebra generated is always finite, implying the termination of our equivalence algorithm; and finally, \emph{complexity}, i.e. the relationship between the number of reachable states and the size of the input expression, which serves as an estimated complexity of our equivalence checking algorithm.
Turns out, all of these questions can be answered by a homomorphism from symbolic Thompson's construction to the symbolic derivatives.

@@

\Cref{thm:hom-thompson-derivative} have several consequences, one of the more obvious one is that we can use the functoriality of the lowering operation to show the semantic equivalence of the start state in the Thompson's construction and the expression in derivative.

@@

\begin{proofEnd}
    By functoriality of lowering~\cref{thm:lowering-functor}, then \(h: S_{e} → D\) is a homomorphism on their lowerings. 
    Because homomorphism preserves semantics (\Cref{thm:hom-preserves-semantics}) \(⟦s^*⟧^{ω}_{S_{e}} = ⟦h(s^*)⟧^{ω}_{D} = ⟦e⟧^{ω}_{D}\), where \(⟦-⟧^{ω}\) is the infinite trace semantics i.e. the unique map into the final GKAT coalgebra \(\mathscr{G}_ω\).

@@

A not so obvious consequence of the homomorphism in~\Cref{thm:hom-thompson-derivative}, is the complexity of the algorithm based on derivatives.
Our bisimulation algorithm (\Cref{alg:symb-bisim}) only explores the principle sub-coalgebra of the start state, i.e. \(s^*\) in the Thompson's construction \(Ŝ_{e}\) or \(e\) in the derivative \(D̂\); thus, deducing an upper bound on the size of the principle sub-coalgebras \(⟨s^*⟩_{Ŝ_{e}}\) and \(⟨e⟩_{D̂}\) are crucial to our complexity analysis.
An upper bound on \(⟨s^*⟩_{Ŝ_{e}}\) is easy to obtain, as the size of \(Ŝ_{e}\), which subsumes the states of \(⟨s^*⟩_{Ŝ_{e}}\), is linear to the size of expression \(e\); therefore \(⟨s^*⟩_{Ŝ_{e}}\) is at most linear to the size of the expression \(e\).
On the other hand the size of \(⟨e⟩_{D̂}\) can, again, be derived from the homomorphism in~\cref{thm:hom-thompson-derivative}.

@@

Although the size of the coalgebra generated by the derivative is smaller than Thompson's construction, the decision procedure based on derivative is not always more efficient than those based on Thompson's construction. 
Crucially, the states in the derivative coalgebra \(D̂\) are expressions, which are more expensive to store; and computing the next transition of the coalgebra can also be more computationally expensive.
Whereas, Thompson's construction only requires inductively going through the expression once to construct the entire coalgebra \(Ŝ_{e}\), and its states can be represented by more efficient constructs, like integers.

>>>>

The final piece of the puzzle is to convert any given expression into an \emph{equivalent} state in some symbolic GKAT coalgebra. 
This goal can be achieved by lifting existing constructions like derivatives and Thompson's construction~\cite{schmid_GuardedKleeneAlgebra_2021,smolka_GuardedKleeneAlgebra_2020} to the symbolic setting.
Then the correctness of the non-symbolic construction can be used to show the correctness of the symbolic construction i.e. we will prove that the lowering as shown in construction~\labelcref{cons:lowering} will yield the conventional derivative.
Other important properties, like finiteness of derivative coalgebra and the correctness of the Thompson's construction can be established using a symbolic GKAT coalgebra homomorphism from the Thompson's construction to the derivative.

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
Basically, a transition is in \(D̂\) if and only if it is derivable by the rules in~\Cref{fig:derivatives-rules}.
These rules are very close to the rules of GKAT derivative by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis for our correctness argument.
\begin{theoremEnd}{theorem}[Correctness]\label{thm:derivative-correctness}
    The lowering of \(D̂\), denoted \(D\), is exactly the derivative defined by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics generated by the derivative coalgebra, \(⟦e⟧ = ⟦e⟧_{D} = ⟦e⟧_{D̂}\).
\end{theoremEnd}
\begin{proofEnd}
    TODO: unfold the statement.
\end{proofEnd}

@@

Another way to construct a coalgebra from a GKAT expression is via Thompson's construction. We lift the original construction to the symbolic setting.
A useful operation for Thompson's construction is the following guard operation, denoted by \(⟨B|\), where \(B\) is a set of boolean expressions: for a transition function \(δ̂\), a accepting function \(ϵ̂\), and a state \(s\),
\begin{align*}
    ⟨B|ϵ̂(s) & ≜ \{b ∧ c ∣ b ∈ B, c ∈ \hat{\epsilon}(s)\}; \\
    ⟨B|δ̂(s) & ≜ \{(b ∧ c, s', p) ∣ b ∈ B, (c, s', p) ∈ \hat{\delta}(s)\}.
\end{align*}
Besides guarding transition and acceptance with different conditions in if-statements and while-loops, the guard operator can also be used to simulate uniform continuation. Specifically, we can use \(⟨ϵ̂(s)|δ(s')\) to connect all the accepting transition of \(s\) to the dynamic \(δ(s')\).

With these definitions in mind, we can define symbolic Thompson's construction inductively as in~\Cref{tab:symb-Thompson-construction}, where we let \((S₁, ϵ̂₁, δ̂₁)\) and \((S₂, ϵ̂₂, δ̂₂)\) to be results of Thompson's construction for \(e₁\) and \(e₂\) respectively.
In this table, \(S₁ + S₂\) denotes the disjoint union of \(S₁\) and \(S₂\), and for any two transition dynamics \(δ₁(s₁): \mathscr{P}(\BExp × S₁ × K)\) and \(δ₂(s₂): \mathscr{P}(\BExp × S₂ × K)\), we can also compose them in parallel as \(δ₁(s₁) + δ₂(s₂)\):
\begin{align*}
    δ₁(s₁) & + δ₂(s₂): \mathscr{P}(\BExp × (S₁ + S₂) × K) \\*
    δ₁(s₁) & + δ₂(s₂) ≜ \\*
        & \{(b, \mathrm{inj}ₗ(s₁'), p) ∣ (b, s₁', p) ∈ δ₁(s₁)\} ∪ \\*
        & \{(b, \mathrm{inj}ᵣ(s₂'), p) ∣ (b, s₂', p) ∈ δ₂(s₂)\},
\end{align*}
where \(\mathrm{inj}ₗ: S₁ → S₁ + S₂\) and \(\mathrm{inj}ᵣ: S₂ → S₁ + S₂\) are the canonical left/right injection of the coproduct.

Our construction deviates from the original construction~\cite{smolka_GuardedKleeneAlgebra_2020} by using a start state \(s^* ∈ S\) instead of a start dynamics (or pseudo-state).
This choice will make the proof of~\Cref{thm:hom-thompson-derivative} slightly easier. 
However, as explained in~\Cref{sec:optimization-implementation}, our implementation uses start dynamics instead of start state to avoid unnecessary lookups.

We would like to explore several desirable theoretical properties of both derivatives and Thompson's construction.
Specifically, \emph{correctness}, i.e. the semantics of the ``start state'' of both constructions preserve the trace semantics of the expression; \emph{finiteness}, i.e. the coalgebra generated is always finite, implying the termination of our equivalence algorithm; and finally, \emph{complexity}, i.e. the relationship between the number of reachable states and the size of the input expression, which serves as an estimated complexity of our equivalence checking algorithm.
It turns out, all of these questions can be answered by a homomorphism from symbolic Thompson's construction to the symbolic derivatives.

@@

\Cref{thm:hom-thompson-derivative} has several consequences, one of the more obvious ones is that we can use the functoriality of the lowering operation to show the semantic equivalence of the start state in the Thompson's construction and the expression in derivative. 
% TODO: what is "in derivative"?

@@

\begin{proofEnd}
    By functoriality of lowering~\Cref{thm:lowering-functor}, then \(h: S_{e} → D\) is a homomorphism on their lowerings. 
    Because homomorphism preserves semantics (\Cref{thm:hom-preserves-semantics}) \(⟦s^*⟧^{ω}_{S_{e}} = ⟦h(s^*)⟧^{ω}_{D} = ⟦e⟧^{ω}_{D}\), where \(⟦-⟧^{ω}\) is the infinite trace semantics i.e. the unique map into the final GKAT coalgebra \(\mathscr{G}_ω\).

@@

A not so obvious consequence of the homomorphism in~\Cref{thm:hom-thompson-derivative} is the complexity of the algorithm based on derivatives.
Our bisimulation algorithm (\Cref{alg:symb-bisim}) only explores the principle sub-coalgebra of the start state, i.e. \(s^*\) in the Thompson's construction \(Ŝ_{e}\) or \(e\) in the derivative \(D̂\); thus, deducing upper bounds on the size of the principle sub-coalgebras \(⟨s^*⟩_{Ŝ_{e}}\) and \(⟨e⟩_{D̂}\) is crucial to our complexity analysis.
An upper bound on \(⟨s^*⟩_{Ŝ_{e}}\) is easy to obtain, as the size of \(Ŝ_{e}\), which subsumes the states of \(⟨s^*⟩_{Ŝ_{e}}\), is linear to the size of expression \(e\); therefore \(⟨s^*⟩_{Ŝ_{e}}\) is at most linear to the size of the expression \(e\).
On the other hand the size of \(⟨e⟩_{D̂}\) can, again, be derived from the homomorphism in~\Cref{thm:hom-thompson-derivative}.

@@

Although the size of the coalgebra generated by the derivative is smaller than Thompson's construction, the states in the derivative coalgebra \(D̂\) are expressions which can be more expensive to store. Computing the next transition of the derivative coalgebra can also take longer. On the other hand, the states of Thompson's construction only requires inductively going through the expression once to construct the entire coalgebra \(Ŝ_{e}\), and its states can be represented by more efficient constructs, like integers. In practice, we utilize techniques such as hashconsing and memoization to address the memory and time usage problems of derivative construction respectively.
